---
title: "Model interpretation and diagnostics"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Model interpretation and diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
```

## Introduction

This vignette covers how to critically interpret `occPlus` model output
and perform basic diagnostic checks. It assumes familiarity with the
introduction vignette (`vignette("occPlus")`).

```{r load}
library(occPlus)
data("sampleresults")
data("sampledata")
sampledata <- data
```

---

## Understanding the model parameters

The `fitmodel` object returned by `runOccPlus()` is a list containing:

| Element | Description |
|---|---|
| `vb_fit` | The `stanfit` object (full MCMC output) |
| `matrix_of_draws` | Posterior samples as a matrix (iterations × parameters) |
| `infos` | Named list of dimensions and labels |
| `X_psi` | Design matrix for occupancy covariates |
| `X_theta` | Design matrix for detection covariates |
| `X_ord` | Design matrix for ordination covariates |

The `infos` element contains:

```{r infos}
str(fitmodel$infos)
```

---

## Posterior summary

The full posterior is stored in `matrix_of_draws`. You can compute
summary statistics for any parameter directly:

```{r posterior-summary}
draws <- fitmodel$matrix_of_draws

# Names of all monitored parameters
head(colnames(draws))

# Posterior median and 95% CI for the first occupancy intercept
beta0 <- draws[, grepl("beta0_psi\\[1\\]", colnames(draws))]
quantile(beta0, probs = c(0.025, 0.5, 0.975))
```

For a tidy overview of all parameters, use `rstan::summary()`:

```{r rstan-summary, eval = FALSE}
rstan::summary(fitmodel$vb_fit)$summary[1:10, ]
```

---

## MCMC diagnostics

Before interpreting results, check that the sampler behaved well.

### HMC diagnostics

```{r hmc-diagnostics, eval = FALSE}
rstan::check_hmc_diagnostics(fitmodel$vb_fit)
```

This prints counts of divergent transitions, saturated tree-depth
proposals, and low Bayesian fraction of missing information (BFMI).
Any non-zero divergence count warrants caution.

### Effective sample size and R-hat

```{r ess-rhat, eval = FALSE}
fit_summary <- rstan::summary(fitmodel$vb_fit)$summary
# Flag parameters with R-hat > 1.05 or ESS < 100
problematic <- fit_summary[fit_summary[, "Rhat"] > 1.05 |
                              fit_summary[, "n_eff"] < 100, ]
head(problematic)
```

An $\hat{R} > 1.05$ for any parameter indicates incomplete convergence.
Running additional chains (set `chains = 3` in `runOccPlus()`) and
checking trace plots is recommended.

### Trace plots

```{r trace, eval = FALSE}
rstan::traceplot(fitmodel$vb_fit,
                 pars = c("mu1", "sigma0", "sigma1"),
                 inc_warmup = FALSE)
```

Well-mixed chains that explore a stationary distribution without
trending indicate good convergence. Sticky or drifting traces suggest
the need for longer warm-up or revised priors.

---

## Interpreting occupancy covariate effects

The occupancy coefficients $\boldsymbol{\beta}^{\psi}_s$ quantify the
effect of site-level covariates on the *logit* occupancy probability,
after accounting for the latent factor structure. A positive coefficient
means the species is more likely to occupy sites with higher values of
that covariate.

```{r occ-effects}
occCov <- returnOccupancyCovariates(fitmodel)
# occCov[iter, covariate, species]
# Posterior median effect of the first covariate on the first species
median(occCov[, 1, 1])
```

Because covariates are automatically **scaled** before fitting, all
coefficients are on a comparable scale and can be compared across
covariates. A change of one posterior-median unit in $\beta^{\psi}$
corresponds to a one-standard-deviation change in the covariate on the
odds-ratio scale.

---

## Interpreting the read intensity model

The parameters $\mu_1$, $\sigma_1$ (true positive reads) and $\sigma_0$
(false positive reads) are shared across species and primers. They
describe the global read-count distributions.

```{r read-params}
mu1    <- median(draws[, "mu1"])
sigma1 <- median(draws[, "sigma1"])
sigma0 <- median(draws[, "sigma0"])

cat(sprintf("True positives:  log(reads+1) ~ N(%.2f, %.2f)\n", mu1, sigma1))
cat(sprintf("False positives: log(reads+1) ~ N(0,    %.2f)\n", sigma0))
cat(sprintf("Implied TP median reads: %.0f\n", exp(mu1) - 1))
```

```{r plot-reads}
plotReadIntensity(fitmodel)
```

A large $\mu_1 / \sigma_0$ ratio and tight $\sigma_1$ indicate that
true positives are well separated from background noise. If $\mu_1$
is small relative to $\sigma_0$, the model may struggle to distinguish
signal from noise; in that case consider using threshold mode
(`threshold` argument) as a robustness check.

---

## Interpreting the latent factor structure

### Factor scores (U)

The factor scores $\mathbf{U} = \mathbf{X}^{\text{ord}}\mathbf{B}^{\text{ord}} + \mathbf{E}$
represent a low-dimensional ordination of the sites in terms of their
species composition after accounting for covariates. Nearby sites in
the factor space tend to host similar species assemblages.

Access the posterior of the factor scores:

```{r factor-scores}
U_draws <- draws[, grepl("^U\\[", colnames(draws))]
n_sites  <- fitmodel$infos$S   # number of sites
d        <- fitmodel$infos$d

# Posterior median factor scores (sites × factors)
U_median <- matrix(apply(U_draws, 2, median),
                   nrow = length(fitmodel$infos$siteNames),
                   ncol = d)
head(U_median)
```

### Factor loadings (LL)

The factor loadings $\boldsymbol{\Lambda}$ quantify how strongly each
species loads on each latent dimension. Large positive loadings on the
same factor indicate species that tend to co-occur above and beyond
what the measured covariates explain.

```{r factor-loadings}
LL_draws  <- draws[, grepl("^LL\\[", colnames(draws))]
LL_median <- matrix(apply(LL_draws, 2, median),
                    nrow = d,
                    ncol = fitmodel$infos$S)
# Rows = factors, columns = species
head(t(LL_median))
```

### Residual correlation matrix

The residual correlation between species $s$ and $s'$ (from the factor
model) can be visualised as described in the introduction vignette. A
useful complementary view is to look at the *magnitude* of the median
correlation across all species pairs:

```{r cor-magnitude, eval = FALSE}
plotCorrelationMatrix(fitmodel)
```

```{r sig-cor}
plotSigElementsCorMatrix(fitmodel)
```

---

## Model selection in detail

### LOO-CV comparison

```{r loo, eval = FALSE}
library(loo)

# Fit candidate models (substitute your data object)
fit_d1 <- runOccPlus(sampledata, d = 1)
fit_d2 <- runOccPlus(sampledata, d = 2)
fit_d3 <- runOccPlus(sampledata, d = 3)

loo1 <- loo(extract_log_lik(fit_d1$vb_fit), save_psis = TRUE)
loo2 <- loo(extract_log_lik(fit_d2$vb_fit), save_psis = TRUE)
loo3 <- loo(extract_log_lik(fit_d3$vb_fit), save_psis = TRUE)

loo_compare(loo1, loo2, loo3)
```

The model with `elpd_diff = 0` provides the best predictive accuracy.
A difference of more than ~4 elpd units between models is generally
considered meaningful.

### Pareto-k diagnostics

```{r pareto-k, eval = FALSE}
plot(loo2)
```

Pareto-$k$ values above 0.7 flag sites that are highly influential.
These are sites where the model fit is poorly constrained by the data
(e.g. very few visits, many missing replicates). Consider collecting
additional samples at these sites or treating their occupancy estimates
with caution.

---

## Posterior predictive checks

A simple posterior predictive check compares the proportion of zeros
predicted by the model to the observed proportion:

```{r ppc, eval = FALSE}
# Extract posterior predictive samples from the Stan fit
# (requires adding generated quantities to the Stan model)
# As a simpler check, compare marginal detection rates:

obs_detection_rate <- mean(sampledata$OTU > 0, na.rm = TRUE)

# Predicted detection rate from posterior collection and FP rates
beta_theta_draws <- draws[, grepl("beta_theta\\[1,", colnames(draws))]
theta_draws <- 1 / (1 + exp(-beta_theta_draws))
pred_detection_rate <- mean(apply(theta_draws, 1, mean))

cat(sprintf("Observed detection rate:  %.3f\n", obs_detection_rate))
cat(sprintf("Predicted detection rate: %.3f\n", pred_detection_rate))
```

---

## Common pitfalls

### Identifiability of the factor model

Factor models have a rotational indeterminacy: $\mathbf{U}\boldsymbol{\Lambda}^\top$
is invariant to orthogonal rotations of $\mathbf{U}$ and
$\boldsymbol{\Lambda}$. `occPlus` imposes an upper-triangular constraint
on the loadings matrix to resolve this. The reparametrisation is applied
automatically in `returnOrdinationCovariatesOutput()`.

### Confounding between occupancy and detection

If site-level covariates also drive detection (e.g. water volume
collected correlates with altitude), including the same variable in
both `occCovariates` and `detCovariates` leads to confounding. Use
domain knowledge to assign covariates appropriately. The `plotReadIntensity()`
output and the magnitude of FP rates are useful diagnostics for
detection-process misspecification.

### Species with very few detections

Species detected at fewer than ~5 sites provide very little information
for estimating occupancy effects. Their posterior distributions for
$\boldsymbol{\beta}^{\psi}_s$ will be wide and dominated by the prior.
Consider aggregating rare species or filtering them before modelling.

---

## Session information

```{r session-info}
sessionInfo()
```
