---
title: "occPlus"
output: 
  rmarkdown::html_vignette:
    fig.width: 7
    fig.height: 4.5
    warning: false
    message: false
vignette: >
  %\VignetteIndexEntry{occPlus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warnings = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  comment = "#>"
)

options(warning = -1)
```

```{r setup, include = FALSE}
library(occPlus)
```
## Overview
The `occPlus` package simplifies the process of modeling occupancy across multiple vertebrate
species using aquatic environmental DNA metabarcoding. It provides tools for occupancy model 
fitting and result visualization in a reproducible workflow.

**Source paper:** High Quality, Granular, Timely, Trustworthy and Efficient Vertebrate Species 
Distribution Data Across a 30,000 km² Protected Area Complex. doi:10.1111/ele.70302

## Data model
- occurrence table: one row per record, required fields (species, lon, lat, date, source).
- meta flags: per-record QC flags and provenance columns.
- derived objects: aggregated summaries and raster/vector outputs.

## Typical usage
1. Ingest raw records (keep originals).
2. Standardize fields and coordinates.
3. Enrich with environmental or detection-effort layers.
4. Aggregate and export reproducible summary artifacts.

### Data preparation

The data object is a list of two elements: `info` and `OTU`. 

`data$info` is a data.frame with $N$ rows, where $N$ is the total
number of replicates analyzed and columns:

- `Site`: name (or index) of the site in which this replicate was collected;
- `Sample`: name (or index) of the sample from which this replicate was analyzed;
- `Primer`: name (or index) of the primer used in this replicate;

`data$OTU` is a matrix of dimension $N \times S$, where $S$ is the number of 
species, where each element is the number of reads of the species in that replicate.
If data are missing, these can be coded as `NA`.

### Running the model
The model can be run using the `runOccPlus` function.

#### Function parameters
- data: see **Data preparation** section above;
- d: number of factors;
- occCovariates: vector of the name of the covariates for the occupancy probabilities. 
- ordCovariates: vector of the name of the covariates for the ordination. 
- detCovariates: vector of the name of the covariates for the detection probabilities. 

**Example:**
```{r, eval = FALSE}
library(occPlus)

# 1. load occurrence data (should be in the form described in the Data preparation section)
dat <- read.csv("path/to/occurrences.csv")

# 2. add spatial precision / grid aggregation
dat_grid <- occPlus::assign_grain(dat_clean, grain = 1000)  # metres
fitmodel  <- runOccPlus(data,
                        d = 2,
                        occCovariates = c("X_psi.1","X_psi.2"),
                        ordCovariates = c("X_ord.1","X_ord.2"),
                        detCovariates = c("X_theta"))

```


The names of the covariates in `occCovariates`, `ordCovariates` and `detCovariates`
should match the column name in data$info.

occPlus does not require a threshold for truncation to fit occupancy models, as
described in the paper. If the user still wants to set a threshold, they can use
the argument *threshold*.

## Model selection

Model selection for the different factors $d$ can be done using the `loo` package.

```{r, eval = FALSE}
library(loo)

fitmodel2  <- runOccPlus(data,
                        d = 3,
                        occCovariates = c("X_psi.1","X_psi.2"),
                        ordCovariates = c("X_ord.1","X_ord.2"),
                        detCovariates = c("X_theta"))

log_lik1 <- extract_log_lik(fitmodel$vb_fit)
log_lik2 <- extract_log_lik(fitmodel2$vb_fit)

loo1 <- loo(log_lik1)
loo1 <- loo(log_lik2)

comparison <- loo_compare(loo1, loo2)
print(comparison)
```

The model with elpd_diff equal to 0 is the one to be preferred

## Output

There are several outputs we can obtain:

### Covariates coefficients

First, we provide three functions for returning occupancy, ordination and detection
covariate coefficients. The output is returned in the form (*niter* x *numberOfCovariates* x *numberOfSpecies*)

```{r}
detCovariatesOutput <- returnDetectionCovariates(fitmodel)
str(detCovariatesOutput)
```

```{r}
ordCovariatesOutput <- returnOrdinationCovariatesOutput(fitmodel)
str(ordCovariatesOutput)
```

```{r}
occCovariatesOutput <- returnOccupancyCovariates(fitmodel)
str(occCovariatesOutput)
```


Next, we provide three functions for plotting occupancy, ordination and detection
covariate coefficients:

```{r}
plotOccupancyCovariates(fitmodel, idx_species = 1:10, covName = "X_psi.1")
```

```{r}
plotOrdinationCovariates(fitmodel, covName = "X_ord.1")
```

```{r}
plotDetectionCovariates(fitmodel, idx_species = 1:10, covName = "X_theta")
```

### Baseline rates

We can return baseline occupancy rates via

```{r}
plotOccupancyRates(fitmodel, idx_species = 1:20)
```

```{r}
plotCollectionRates(fitmodel, idx_species = 1:20)
```


```{r}
plotDetectionRates(fitmodel, idx_species = 1:20)
```

And false positives rates at the first stage (field stage):

```{r}
plotStage1FPRates(fitmodel, idx_species = 1:20)
```

And at the second stage (lab stage):

```{r}
plotStage2FPRates(fitmodel, idx_species = 1:20)
```

### Occupancy rates

To compute occupancy probabilities at each site and for each species, we use
the following function. 

```{r, eval = T}
occProbs <- computeOccupancyProbs(fitmodel)
str(occProbs)
```


The output is stored in the form (*iterations* x *sites* x *species*).




### Read intensity

`occPlus` estimates a distribution of reads for true and false positives. We can
plot the two distributions via:

```{r, fig.width=7, fig.height=4.5, warnings = FALSE}
plotReadIntensity(fitmodel)
```

### Covariance matrix

We can plot the posterior median of the residual correlation matrix as

```{r}
plotCorrelationMatrix(fitmodel)
```


We can show only the positively and negatively significant elements (positive in
red and negative in blue)

```{r}
plotSigElementsCorMatrix(fitmodel)
```

We note that this is the \textit{residual} covariance matrix, \textbf{after} 
accounting for occupancy covariates but accounting for ordination covariates.

#### References
- Paper: High Quality, Granular, Timely, Trustworthy and Efficient Vertebrate Species Distribution Data Across a 30,000 km² Protected Area Complex. doi:10.1111/ele.70302
