---
title: "Introduction to occPlus"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 4.5
    warning: false
    message: false
vignette: >
  %\VignetteIndexEntry{Introduction to occPlus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)

options(warning = -1)
```

## Overview

`occPlus` fits a Bayesian hierarchical occupancy model designed for
**multi-species, multi-marker eDNA metabarcoding data**. The model
simultaneously infers species occupancy probabilities, sample-level
collection (detection) probabilities, and primer-level true- and
false-positive read rates—all while accounting for residual species
co-occurrence via a latent factor structure.

The method is described in full in:

> Ji Y, Diana A, Li X, Matechou E, Griffin JE, Liu S, *et al.* (2025).
> High Quality, Granular, Timely, Trustworthy and Efficient Vertebrate
> Species Distribution Data Across a 30,000 km² Protected Area Complex.
> *Ecology Letters*, 28(12).
> <https://doi.org/10.1111/ele.70302>

### When to use `occPlus`

| Feature | Details |
|---|---|
| Data type | eDNA metabarcoding read counts (or binary detections) |
| Survey design | Multiple sites, samples per site, PCR replicates per sample, one or more primers |
| Species | Any number; modelled jointly |
| Inference engine | Bayesian MCMC |
| False-positive modelling | Two-stage: field (sample) and lab (PCR replicate) |
| Covariate support | Occupancy, detection, and ordination (latent factor) covariates |

---

```{r setup, include = FALSE}
library(occPlus)
```

The model can be run using the `runOccPlus` function.

The arguments of the function are:

- data: which we explain below;
- d: number of factors;
- occCovariates: vector of the name of the covariates for the occupancy probabilities.
- ordCovariates: vector of the name of the covariates for the ordination.
- collCovariates: vector of the name of the covariates for the detection probabilities.

## Data preparation

The data object is a list of two elements: `info` and `OTU`.

`data$info` is a data.frame with $N$ rows, where $N$ is the total
number of replicates analyzed and columns:

- `Site`: name (or index) of the site in which this replicate was collected;
- `Sample`: name (or index) of the sample from which this replicate was analyzed;
- `Primer`: name (or index) of the primer used in this replicate;


```{r}
head(data$info)
```

`data$OTU` is a matrix of dimension $N \times S$, where $S$ is the number of
species, where each element is the number of reads of the species in that replicate.
If data are missing, these can be coded as `NA`.

```{r}
head(data$OTU)
```

## Model fitting

Next, we can run the model by typing

```{r, eval = FALSE}
fitmodel  <- runOccPlus(data,
                        d = 2,
                        occCovariates = c("X_psi.1","X_psi.2"),  # predictors of occupancy (ψ)
                        ordCovariates = c("X_ord.1","X_ord.2"),  # predictors of latent factors (U)
                        collCovariates = c("X_theta"))           # predictors of collection probability (θ)

```


The names of the covariates in `occCovariates`, `ordCovariates` and `collCovariates`
should match the column name in data$info.

occPlus does not require a threshold for truncation to fit occupancy models, as
described in the paper. If the user still wants to set a threshold, they can use
the argument *threshold*.

## Model selection

Model selection for the different factors $d$ can be done using the `loo` package.

```{r, eval = FALSE}
library(loo)

fitmodel2  <- runOccPlus(data,
                        d = 3,
                        occCovariates = c("X_psi.1","X_psi.2"),
                        ordCovariates = c("X_ord.1","X_ord.2"),
                        collCovariates = c("X_theta"))

log_lik1 <- extract_log_lik(fitmodel$vb_fit)
log_lik2 <- extract_log_lik(fitmodel2$vb_fit)

loo1 <- loo(log_lik1)
loo1 <- loo(log_lik2)

comparison <- loo_compare(loo1, loo2)
print(comparison)
```

The model with elpd_diff equal to 0 is the one to be preferred

## Output

There are several outputs we can obtain:

### Covariates coefficients

First, we provide three functions for returning occupancy, ordination and detection
covariate coefficients. The output is returned in the form (*niter* x *numberOfCovariates* x *numberOfSpecies*)

```{r plot-det-covs, eval = FALSE}
collCovariatesOutput <- returnDetectionCovariates(fitmodel)
str(collCovariatesOutput)
```

```{r return-ord-covs, eval = FALSE}
ordCovariatesOutput <- returnOrdinationCovariatesOutput(fitmodel)
str(ordCovariatesOutput)
```

```{r plot-occ-covs, eval = FALSE}
occCovariatesOutput <- returnOccupancyCovariates(fitmodel)
str(occCovariatesOutput)
```


Next, we provide three functions for plotting occupancy, ordination and detection
covariate coefficients:

```{r plot-occ-covs1, eval = FALSE}
plotOccupancyCovariates(fitmodel, idx_species = 1:10, covName = "X_psi.1")
```

```{r return-ord-covs1, eval = FALSE}
plotOrdinationCovariates(fitmodel, covName = "X_ord.1")
```

```{r plot-det-covs1, eval = FALSE}
plotDetectionCovariates(fitmodel, idx_species = 1:10, covName = "X_theta")
```

### Baseline rates

We can return baseline occupancy rates via

```{r plot-occ-covs2, eval = FALSE}
plotOccupancyRates(fitmodel, idx_species = 1:20)
```

```{r plot-collection, eval = FALSE}
plotCollectionRates(fitmodel, idx_species = 1:20)
```


```{r plot-detection, eval = FALSE}
plotDetectionRates(fitmodel, idx_species = 1:20)
```

And false positives rates at the first stage (field stage):

```{r plot-stage1-fp, eval = FALSE}
plotStage1FPRates(fitmodel, idx_species = 1:20)
```

And at the second stage (lab stage):

```{r plot-stage2-fp, eval = FALSE}
plotStage2FPRates(fitmodel, idx_species = 1:20)
```

### Occupancy rates

To compute occupancy probabilities at each site and for each species, we use
the following function.

```{r occ-probs, eval = FALSE}
occProbs <- computeOccupancyProbs(fitmodel)
str(occProbs)
```


The output is stored in the form (*iterations* x *sites* x *species*).




### Read intensity

`occPlus` estimates a distribution of reads for true and false positives. We can
plot the two distributions via:

```{r plot-reads, fig.width=7, fig.height=4.5, warnings = FALSE, eval = FALSE}
plotReadIntensity(fitmodel)
```

### Covariance matrix

We can plot the posterior median of the residual correlation matrix as

```{r plot-cor, eval = FALSE}
plotCorrelationMatrix(fitmodel)
```


We can show only the positively and negatively significant elements (positive in
red and negative in blue)

```{r plot-sig-cor, eval = FALSE}
plotSigElementsCorMatrix(fitmodel)
```

We note that this is the \textit{residual} covariance matrix, \textbf{after}
accounting for occupancy covariates but accounting for ordination covariates.

---

## Citation

If you use `occPlus` in published work, please cite:

```
Ji Y, Diana A, Li X, Matechou E, Griffin JE, Liu S, Luo M, Wu C,
Bai R, Yao C, Yin T, Dong F, Wu F, Wang K, Yu Z, Chen X, Jiang X,
Che J, Yu DW, Popescu VD (2025). High Quality, Granular, Timely,
Trustworthy and Efficient Vertebrate Species Distribution Data Across
a 30,000 km² Protected Area Complex. Ecology Letters, 28(12).
doi:10.1111/ele.70302
```

---

## Session information

```{r session-info}
sessionInfo()
```
