---
title: "Introduction to occPlus"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Introduction to occPlus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
```

## Overview

`occPlus` fits a Bayesian hierarchical occupancy model designed for
**multi-species, multi-marker eDNA metabarcoding data**. The model
simultaneously infers species occupancy probabilities, sample-level
collection (detection) probabilities, and primer-level true- and
false-positive read rates—all while accounting for residual species
co-occurrence via a latent factor structure.

The method is described in full in:

> Ji Y, Diana A, Li X, Matechou E, Griffin JE, Liu S, *et al.* (2025).
> High Quality, Granular, Timely, Trustworthy and Efficient Vertebrate
> Species Distribution Data Across a 30,000 km² Protected Area Complex.
> *Ecology Letters*, 28(12).
> <https://doi.org/10.1111/ele.70302>

### When to use `occPlus`

| Feature | Details |
|---|---|
| Data type | eDNA metabarcoding read counts (or binary detections) |
| Survey design | Multiple sites, samples per site, PCR replicates per sample, one or more primers |
| Species | Any number; modelled jointly |
| Inference engine | Bayesian MCMC via Stan (`rstan`) |
| False-positive modelling | Two-stage: field (sample) and lab (PCR replicate) |
| Covariate support | Occupancy, detection, and ordination (latent factor) covariates |

---

## Installation

```{r install, eval = FALSE}
# Install from GitHub (development version)
# install.packages("remotes")
remotes::install_github("AlexDiana/occPlus")
```

`occPlus` links to [Stan](https://mc-stan.org/) through `rstan`. Make
sure a working C++ toolchain is present before installation; see the
[RStan Getting Started](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started)
guide for platform-specific instructions.

---

## The statistical model

The model operates on a three-level hierarchy that mirrors the eDNA
sampling process:

```
Site  (occupancy z_is)
  └─ Sample  (collection w_ims)
       └─ PCR replicate  (read count y_imkls)
```

**Level 1 — Occupancy.** Species $s$ occupies site $i$ with probability
$\psi_{is}$, modelled on the logit scale as

$$
\text{logit}(\psi_{is}) = \beta^{\psi}_{0s} + \mathbf{x}_i^{\top}\boldsymbol{\beta}^{\psi}_s
  + \mathbf{u}_i^{\top}\boldsymbol{\lambda}_s,
$$

where $\mathbf{u}_i = \mathbf{x}_i^{\text{ord}}\mathbf{B}^{\text{ord}} + \mathbf{e}_i$
are latent factor scores that capture residual species co-occurrence
after accounting for measured covariates.

**Level 2 — Collection.** Within an occupied site, sample $m$ captures
eDNA of species $s$ with probability $\theta_{ims}$ (from unoccupied
sites, with the lower baseline rate $\theta_{0s}$):

$$
\text{logit}(\theta_{ims}) = \mathbf{x}_{im}^{\top}\boldsymbol{\beta}^{\theta}_s.
$$

**Level 3 — Read intensity.** Each PCR replicate yields reads modelled
as a mixture: true-positive reads follow
$\log(y+1) \sim \mathcal{N}(\mu_1, \sigma_1^2)$ and false-positive
reads follow a zero-inflated normal centred at zero. The primer-specific
true-positive rate is $p_{ls}$ and false-positive rate is $q_{ls}$.

This design allows `occPlus` to distinguish genuine detections from
contamination or cross-amplification artefacts **without requiring a
fixed read-count threshold**.

---

## Data preparation

The input to `runOccPlus()` is a named list with two elements.

### `data$info`

A `data.frame` with **one row per PCR replicate** and at least these
three columns (additional columns serve as potential covariates):

| Column | Description |
|---|---|
| `Site` | Site identifier (character or factor) |
| `Sample` | Sample identifier, unique within a site |
| `Primer` | Primer / marker identifier |

### `data$OTU`

A numeric matrix of dimensions $N_\text{replicates} \times S$, where
each cell is the raw read count of species $s$ in that replicate. Rows
must be in the same order as `data$info`. Missing replicates should be
coded as `NA`.

### Inspecting the example dataset

```{r check-data, eval = TRUE}
library(occPlus)

# The package ships with an illustrative example dataset
data("sampledata")

# Inspect the metadata
head(sampledata$info)

# Inspect the OTU table (first 5 species)
sampledata$OTU[1:6, 1:5]
```

Key checks before fitting:

- `Site`, `Sample`, and `Primer` must contain no `NA` values.
- Covariate columns used in the model must be present in `data$info`.
- OTU matrix rows must align with `data$info` rows.

---

## Fitting the model

### Minimal call

```{r fit-minimal, eval = FALSE}
fitmodel <- runOccPlus(
  data = sampledata,
  d    = 2        # number of latent factors
)
```

### Adding covariates

Covariates are specified by supplying column names from `data$info`.
`occPlus` automatically scales continuous predictors and expands
categorical ones into dummy variables.

```{r fit-covariates, eval = FALSE}
fitmodel <- runOccPlus(
  data            = sampledata,
  d               = 2,
  occCovariates   = c("altitude"),          # predictors of occupancy (ψ)
  ordCovariates   = c("altitude", "NDVI"),  # predictors of latent factors (U)
  detCovariates   = c("input_volume")       # predictors of collection probability (θ)
)
```

The names in `occCovariates`, `ordCovariates`, and `detCovariates`
must exactly match column names in `data$info`.

### Binary threshold mode

To binarise read data before fitting (threshold-based occupancy
modelling), supply a numeric cutoff. Any replicate with reads
$\geq$ `threshold` is treated as a detection.

```{r fit-binary, eval = FALSE}
fitmodel_bin <- runOccPlus(
  data      = sampledata,
  d         = 2,
  threshold = 10   # reads >= 10 treated as detection
)
```

### Key arguments

| Argument | Default | Description |
|---|---|---|
| `data` | — | Named list with `$info` and `$OTU` |
| `d` | — | Number of latent factors (≥ 0) |
| `threshold` | `FALSE` | Read-count threshold; `FALSE` uses continuous read model |
| `occCovariates` | `c()` | Occupancy covariate names |
| `ordCovariates` | `c()` | Ordination covariate names |
| `detCovariates` | `c()` | Detection covariate names |
| `numSamples` | `350` | Number of posterior samples retained |

The Stan sampler runs for 2 000 iterations (1 000 warm-up) with a
single chain (`adapt_delta = 0.99`, `max_treedepth = 20`).

---

## Exploring model output

All plotting and extraction functions take the `fitmodel` object
returned by `runOccPlus()` as their first argument. Interval-based
plots display **95% posterior credible intervals** by default.

For the examples below we load the pre-computed results shipped with
the package:

```{r load-results}
data("sampleresults")
fitmodel <- sampleresults
```

### Occupancy probabilities

`computeOccupancyProbs()` returns a three-dimensional array of
dimensions (*iterations* × *sites* × *species*), containing the full
posterior distribution of $\psi_{is}$.

```{r occ-probs}
occProbs <- computeOccupancyProbs(fitmodel)
str(occProbs)
```

Visualise site-averaged occupancy rates across species:

```{r plot-occ-rates}
plotOccupancyRates(fitmodel, idx_species = 1:20)
```

### Covariate effects

#### Occupancy covariates

Credible intervals that exclude zero (red horizontal line) indicate a
statistically meaningful effect.

```{r return-occ-covs}
occCovOutput <- returnOccupancyCovariates(fitmodel)
# Array dimensions: [iterations, covariates, species]
str(occCovOutput)
```

```{r plot-occ-covs, eval = FALSE}
# Replace "altitude" with the name of your occupancy covariate
plotOccupancyCovariates(fitmodel, covName = "altitude", idx_species = 1:20)
```

#### Ordination (latent factor) covariates

```{r return-ord-covs}
ordCovOutput <- returnOrdinationCovariatesOutput(fitmodel)
str(ordCovOutput)
```

```{r plot-ord-covs, eval = FALSE}
plotOrdinationCovariates(fitmodel, covName = "altitude")
```

#### Detection (collection) covariates

```{r return-det-covs}
detCovOutput <- returnDetectionCovariates(fitmodel)
str(detCovOutput)
```

```{r plot-det-covs, eval = FALSE}
plotDetectionCovariates(fitmodel, covName = "input_volume", idx_species = 1:20)
```

### Baseline rates

**Collection rate** — probability that an occupied site yields a
positive sample, at average covariate values:

```{r plot-collection}
plotCollectionRates(fitmodel, idx_species = 1:20)
```

**Lab-stage true-positive rate** ($p_{ls}$) — probability that a
positive replicate produces reads given a positive sample, per primer:

```{r plot-detection}
plotDetectionRates(fitmodel, idx_species = 1:20)
```

### False-positive rates

**Field-stage FP rate** ($\theta_{0s}$) — probability of a positive
sample at an *unoccupied* site:

```{r plot-stage1-fp}
plotStage1FPRates(fitmodel, idx_species = 1:20)
```

**Lab-stage FP rate** ($q_{ls}$) — false-positive read rate per PCR
replicate per primer:

```{r plot-stage2-fp}
plotStage2FPRates(fitmodel, idx_species = 1:20)
```

### Read intensity distributions

The model estimates the log-read distributions for true positives
($\mathcal{N}(\mu_1, \sigma_1^2)$) and false positives
(zero-inflated normal). The shaded bands show 80% and 95% credible
envelopes.

```{r plot-reads}
plotReadIntensity(fitmodel)
```

Good separation between the true-positive (right) and false-positive
(left) distributions indicates high discriminability. Substantial
overlap suggests adding more PCR replicates or switching to threshold
mode.

### Residual species co-occurrence

After accounting for measured covariates, the latent factor model
captures residual correlations in species occupancy. Positive values
(red) indicate co-occurring pairs; negative values (blue) indicate
avoidance.

Note: this is the *residual* correlation matrix—it reflects associations
**after** occupancy covariates have been accounted for, but **including**
any shared environmental gradients captured by the ordination factors.

```{r plot-cor}
plotCorrelationMatrix(fitmodel)
```

Highlight only credible non-zero correlations (95% CI excluding zero):

```{r plot-sig-cor}
plotSigElementsCorMatrix(fitmodel)
```

---

## Model selection: choosing `d`

The number of latent factors `d` is the primary structural
hyperparameter. Use leave-one-out cross-validation via the
[`loo`](https://CRAN.R-project.org/package=loo) package to compare
candidate models.

```{r model-selection, eval = FALSE}
library(loo)

fitmodel_d0 <- runOccPlus(sampledata, d = 0)  # standard multi-species model
fitmodel_d1 <- runOccPlus(sampledata, d = 1)
fitmodel_d2 <- runOccPlus(sampledata, d = 2)
fitmodel_d3 <- runOccPlus(sampledata, d = 3)

loo0 <- loo(extract_log_lik(fitmodel_d0$vb_fit))
loo1 <- loo(extract_log_lik(fitmodel_d1$vb_fit))
loo2 <- loo(extract_log_lik(fitmodel_d2$vb_fit))
loo3 <- loo(extract_log_lik(fitmodel_d3$vb_fit))

loo_compare(loo0, loo1, loo2, loo3)
```

The model with `elpd_diff = 0` is preferred. Setting `d = 0` fits
a standard multi-species occupancy model with no latent structure and
provides a useful null baseline. In practice, `d = 2` or `d = 3` is
sufficient for most assemblages.

---

## Tips and common issues

**Stan compilation.** The Stan model is compiled at the first call to
`runOccPlus()` within a session. Subsequent calls within the same
session reuse the compiled model and are considerably faster.

**Convergence checks.** Run
`rstan::check_hmc_diagnostics(fitmodel$vb_fit)` after fitting.
Divergent transitions or low effective sample sizes indicate potential
convergence problems. Try longer chains or a simpler covariate
structure.

**Categorical covariates.** Factor columns in `data$info` are
automatically expanded into dummy variables. The resulting column names
(e.g. `"typeW"`) are the names to pass to plotting functions; use
`colnames(fitmodel$X_psi)` to inspect them.

**Missing data.** Code missing read counts in `data$OTU` as `NA`.
The model excludes missing replicates from the likelihood automatically.

**Large datasets.** For datasets with many sites, species, and
replicates, reduce `numSamples` during prototyping and increase it
for the final analysis.

---

## Citation

If you use `occPlus` in published work, please cite:

```
Ji Y, Diana A, Li X, Matechou E, Griffin JE, Liu S, Luo M, Wu C,
Bai R, Yao C, Yin T, Dong F, Wu F, Wang K, Yu Z, Chen X, Jiang X,
Che J, Yu DW, Popescu VD (2025). High Quality, Granular, Timely,
Trustworthy and Efficient Vertebrate Species Distribution Data Across
a 30,000 km² Protected Area Complex. Ecology Letters, 28(12).
doi:10.1111/ele.70302
```

---

## Session information

```{r session-info}
sessionInfo()
```
